Altering Backward Pass Gradients to improve Convergence

### 2022-11-15 0100.tex
PAKDD Mod1

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{framed}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{url}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{booktabs}
% \usepackage{amsthm}
\usepackage{pifont}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{color}
\graphicspath{{images/}}
\renewcommand\UrlFont{\color{blue}\rmfamily}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\NewDocumentCommand{\codeword}{v}{\texttt{\textcolor{orange}{#1}}}
\newcommand{\cmark}{\ding{51}} \newcommand{\xmark}{\ding{55}}
\definecolor{newcolor}{rgb}{.8,.349,.1}
\definecolor{codegreen}{rgb}{0.7,0.7,0.7}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{1,1,1}
\lstdefinestyle{mystyle}{ backgroundcolor=\color{backcolour},   
commentstyle=\color{codegreen}, keywordstyle=\color{magenta},
numberstyle=\tiny\color{codegray}, stringstyle=\color{codepurple},
basicstyle=\ttfamily\footnotesize, breakatwhitespace=false,         
breaklines=true,                 
captionpos=b,                    
keepspaces=true,                 
numbersep=5pt,                  
showspaces=false,                
showstringspaces=false, showtabs=false,                  
tabsize=2 }
\lstset{style=mystyle}

\begin{document}
\title{Altering Backward Pass Gradients to improve Convergence}
\author{Bishshoy Das\inst{1}\orcidID{0000-0002-6525-4015}\and Milton Mondal\inst{1}\orcidID{xxxx-xxxx-xxxx-xxxx}\and Brejesh Lall\inst{3}\orcidID{xxxx-xxxx-xxxx-xxxx}\and Shiv Dutt Joshi\inst{3}\orcidID{xxxx-xxxx-xxxx-xxxx}\and Sumantra Dutta Roy\inst{3}\orcidID{xxxx-xxxx-xxxx-xxxx}}
\authorrunning{B. Das et al.}
\institute{Indian Institute of Technology Delhi, New Delhi - 110016, India \\ \email{\{bishshoy.das, milton.mondal, brejesh, sdjoshi, sumantra\}@ee.iitd.ac.in}}
\maketitle

@@

\subsection{Properties of PGT}
\label{sec:pgt_prop}

@@

\textit{Proof.} The proof follows directly from the definition of $p_i'$. We notice that
if $p_i' < p_i$, we get:

\begin{proof} \begin{alignat}{2} &\ \ \frac{p_i^{\alpha }}{\sum
_{j=1}^C p_j^{\alpha}} < p_i \\ \Rightarrow &\ p_i^{\alpha -1}<\sum _{j=1}^C p_j^{\alpha
} \\ \Rightarrow &\ p_i > \Big(\sum _{j=1}^C p_j^{\alpha}\Big)^{\frac{1}{\alpha-1}}
\end{alignat} \end{proof}

@@

@@

Since we have established that values of $p_i$ which are greater than the stationary
threshold decreases and move down towards the stationary threshold, and values in $p_i$
lower than the stationary threshold moves up towards the stationary threshold, this
transformation makes the distribution more uniform (i.e. it smooths out the actual
distribution) as $\alpha$ is decreased from $1$ and down towards $0$. This final
observation we prove in the following theorem.

@@

If $a_1,\dots ,a_n$ and $b_1,\dots ,b_n$ are non-negative numbers, then using the log
sum inequality, \\ we get $\sum _{j=1}^n a_j \log \left(\frac{a_j}{b_j}\right)\geq
\left(\sum _{j=1}^n a_j\right) \log \left(\frac{\sum _{j=1}^n a_j}{\sum _{j=1}^n
b_j}\right)$. Setting $a_j=p_j^\alpha$ and $b_j=1$, we get the following lower bound
\begin{equation} \sum _{j=1}^C p_j^{\alpha } \log
\left(p_j^{\alpha }\right)\geq \left(\sum _{j=1}^C
p_j^{\alpha}\right) \log \left(\frac{1}{C} \sum _{j=1}^C
p_j^{\alpha}\right) \label{lower_bound1} \end{equation}

@@

\begin{proof} \begin{equation} \frac{\partial}{\partial \alpha
}\left(\sum _{j=1}^c p_j^{\alpha}\right){}^{\frac{1}{\alpha -1}}\ \ \geq \ \ 0
\end{equation} \end{proof}

@@

Now, we demonstrate how PGT restricts the magnitude of the directional derivative for
each logit. For a training example, a neural network can assign the highest probability
to either (i) a wrong class or to the (ii) correct class.

@@

\begin{table}[!t]
\centering
\caption{ Results and comparison table for CNN networks trained on ImageNet-1K (sorted
in descending order w.r.t. the last column denoting test accuracy improvements over
baseline). Best training and test accuracies are highlighted in red and blue
respectively. Accuracy differences are highlighted in yellow. In all experiments, we
observe that PGT positively improves both training and test accuracies. We also observe
that deeper networks such as SE-ResNet-50 and ResNet-50 accrue higher gains over the
baselines as compared to shallow networks like ResNet-18. }
\label{tab:imagenet_table}
\begin{tabular}{cccccccc}
\multirow{2}{*}{\textbf{Model}} & \multirow{2}{*}{\textbf{Scheduler}} &
\multirow{2}{*}{\textbf{Method}} & \textbf{PGT} & \textbf{\hspace{-0.65cm} Train}
& \textbf{Train} & \textbf{\hspace{-0.65cm} Test} & \textbf{Test} \\
& & & \textbf{($\alpha$)} & \textbf{Acc. (\%)} & \textbf{Diff (\%)} &
\textbf{Acc. (\%)} & \textbf{Diff (\%)} \\
\midrule
\multirow{2}{*}{SE-ResNet-50} & \multirow{2}{*}{Cosine} & Baseline & - & 81.5 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.97}}} & 77.218 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.734}}} \\
& & PGT & 0.3 & \textcolor{red}{\textbf{82.47}} & & \textcolor{blue}{\textbf{77.952}} &
\\
\midrule
\multirow{2}{*}{ResNet-50} & \multirow{2}{*}{Cosine} & Baseline & - & 79.18 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.5}}} & 76.56 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.656}}} \\
& & PGT & 0.05 & \textcolor{red}{\textbf{79.68}} & & \textcolor{blue}{\textbf{77.216}} &
\\
\midrule
\multirow{2}{*}{ResNet-50} & \multirow{2}{*}{Step} & Baseline & - & 78.99 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.57}}} & 75.97 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.524}}} \\
& & PGT & 0.05 & \textcolor{red}{\textbf{79.56}} & & \textcolor{blue}{\textbf{76.494}} &
\\
\midrule
\multirow{2}{*}{ResNet-101} & \multirow{2}{*}{Cosine} & Baseline & - & 82.29 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.81}}} & 77.896 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.362}}} \\
& & PGT & 0.3 & \textcolor{red}{\textbf{83.1}} & & \textcolor{blue}{\textbf{78.258}} &
\\
\midrule
\multirow{2}{*}{SE-ResNet-18} & \multirow{2}{*}{Cosine} & Baseline & - & 71.42 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.18}}} & 71.09 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.346}}} \\
& & PGT & 0.25 & \textcolor{red}{\textbf{71.6}} & & \textcolor{blue}{\textbf{71.436}} &
\\
\midrule
\multirow{2}{*}{ResNet-18} & \multirow{2}{*}{Step} & Baseline & - & 69.95 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.35}}} & 69.704 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.14}}} \\
& & PGT & 0.25 & \textcolor{red}{\textbf{70.3}} & & \textcolor{blue}{\textbf{69.844}} &
\\
\midrule
\multirow{2}{*}{ResNet-18} & \multirow{2}{*}{Cosine} & Baseline & - & 70.38 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.15}}} & 70.208 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.09}}} \\
& & PGT & 0.25 & \textcolor{red}{\textbf{70.53}} & & \textcolor{blue}{\textbf{70.298}} &
\\
\end{tabular}
\end{table}

In our experiments with Squeeze-and-Excitation variant of ResNet-50 i.e.
SE-ResNet-50\cite{hu2018squeeze}, we observe significant improvements; a $0.97\%$ boost
in training accuracy and a $0.734\%$ increase in test accuracy. In our experiments with
ResNet-50, we find an increase of $0.5\%$ and $0.656\%$ performance enhancement over the
cosine scheduler baseline for training and testing respectively. The corresponding
improvement over the step scheduler baseline for ResNet-50 is $0.57\%$ (training) and
$0.524\%$ (testing). ResNet-101 sees a higher improvement in training fit, $0.81\%$ to
be exact, while the improvement over the test set is $0.362\%$. Smaller networks such as
SE-ResNet-18 and ResNet-18 sees accuracy boosts which are smaller but nevertheless
positive. Therefore we conclude that with PGT, the consistent improvements in training
accuracies across all cases is because the networks train better and arrive at better
optimas during convergence. Per epoch training and test accuracy plots of ResNet-18 and
ResNet-50 (both with and without PGT) are shown in Fig. \ref{fig:metrics}(a, b). For
better visualization of the top end of the plots, we scale both axes in log scale.
Because both training and test accuracies improve, it indicates that PowerGrad Transform
allows the network to learn better representations\footnote{Reproducible code, training
recipes for the experiments, pretrained checkpoints and training logs are provided at:
\url{ https://github.com/bishshoy/power-grad-transform }. Code is adapted from
\cite{rw2019timm}.}.

This section empirically examines some of the issues that occur when networks are
trained without normalization layers and we provide insight into the effects of
PowerGrad transform on such networks. We use ResNet-18 \cite{he2016deep} as the
foundation model for all empirical trials owing to its popularity among deep learning
practitioners and the relative simplicity with which it can be trained on a big dataset
such as ImageNet-1K \cite{russakovsky2015imagenet}. We found deeper networks (such as
ResNet-34 and ResNet-50 models) are impossible to train without BatchNormalization
because of increased depth. So in order to conduct our empirical experiments, we solely
focus on the ResNet-18 architecture. For our experiments, we designate different layers
with their corresponding layer indices. We provide a detailed layer-wise index list of
ResNet-18 in the supplementary section.

We concentrate on the following metric: the per-filter L2-norm of each layer's weight
tensor. Throughout the training process, we monitor variations in the per-filter weight
norm. Because a full plot of each layer's data would use an inordinate amount of space,
we focus on a few critical layers' plots. The supplementary section contains full plots
for all layer for various settings. The norm of the weights of the $11^\textit{th}$
convolutional layer is shown in Fig. \ref{fig:norm_plots}(a). Layer $11$ includes $256$
filters. Each colour on the figure represents the iteration-wise evolution of a
particular filter throughout the course of training. The critical point to note is that
some filters achieve a norm of zero during training. We refer to this event as `Zeroing
Out', and occurs when a channel of a weight tensor gets fully filled with zeros. If all
the coefficients of the filter becomes zero, only then the L2 norm of that filter
becomes zero. If we observe zero filter norm for a filter, then that filter does not
contribute at all to determine the input-output relationship of a dataset, as the
feature tensor it produces is also an all-zero tensor. According to our findings, when a
filter tensor becomes zeroed out, it does not recover with further training. We observe
both effects \textbf{1)} zeroed out filter \textbf{2)} zeroed out feature.

\begin{table}[!t]
\centering
\caption{ Results for non-normalized ResNets on ImageNet-1K. Each experiment is
conducted with a 100 epoch budget using the cosine decay scheduler. Best training and
test accuracies are highlighted in red and blue respectively. Top differences in
training and test accuracies are marked in yellow. The baselines in each segment is
taken as the reference point for all `Diff' accuracies in that segment. The first three
segments highlight comparison for experiments where the methods are applied
individually, while the last segment denotes performance metrics for combinations of
different methods (GC + PGT, AGC + PGT). We see that at any batch size, with PGT, the
performance of non-normalized version of ResNet-18 improves significantly over the
baseline owing to the improvements in training as observed in Fig. \ref{fig:agc_pgt}. }
\label{tab:wobn_table}
\begin{tabular}{cccccccc}
\multirow{2}{*}{\textbf{Model}} & \multirow{2}{*}{\textbf{Batch Size}} &
\multirow{2}{*}{\textbf{Method}} & \textbf{PGT} & \textbf{\hspace{-0.65cm} Train}
& \textbf{Train} & \textbf{\hspace{-0.65cm} Test} & \textbf{Test} \\
& & & \textbf{($\alpha$)} & \textbf{Acc. (\%)} & \textbf{Diff (\%)} &
\textbf{Acc. (\%)} & \textbf{Diff (\%)} \\
\midrule
ResNet-18 & 1024 & Baseline & - & 66.27 & - & 64.816 & - \\
(non-BN) & 1024 & PGT & 0.92 & \textcolor{red}{\textbf{66.62}} &
\textcolor{olive}{\textbf{+0.35}} & \textcolor{blue}{\textbf{65.498}} &
\textcolor{olive}{\textbf{+0.682}} \\
\cmidrule{2-8}
& 512 & Baseline & - & 68.02 & - & 66.552 & - \\
& 512 & PGT & 0.25 & \textcolor{red}{\textbf{69.5}} & \textcolor{olive}{\textbf{+1.48}}
& \textcolor{blue}{\textbf{67.236}} & \textcolor{olive}{\textbf{+0.684}} \\
\cmidrule{2-8}
& 256 & Baseline & - & 68.86 & - & 66.796 & - \\
& 256 & GC & - & 69.04 & +0.18 & 67.064 & +0.268 \\
& 256 & AGC & - & 69.06 & +0.2 & 67.298 & +0.502 \\
& 256 & PGT & 0.25 & \textcolor{red}{\textbf{69.97}} & \textcolor{olive}{\textbf{+1.11}}
& \textcolor{blue}{\textbf{67.814}} & \textcolor{olive}{\textbf{+1.018}} \\
\cmidrule{2-8}
& 256 & Baseline & - & 68.86 & - & 66.796 & - \\
& 256 & GC+PGT & 0.25 & 68.67 & -0.19 & 67.088 & +0.292 \\
& 256 & AGC+PGT & 0.25 & \textcolor{red}{\textbf{70.92}} &
\textcolor{olive}{\textbf{+2.06}} & \textcolor{blue}{\textbf{68.856}} &
\textcolor{olive}{\textbf{+2.06}} \\
\end{tabular}
\end{table}

\begin{figure}[t]
\centering
\begin{subfigure}{0.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{logit_vs_alpha}
\caption{\\Logit vs. $\alpha$}
\end{subfigure}%
\begin{subfigure}{0.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{loss_vs_alpha}
\caption{\\Cross-entropy loss vs. $\alpha$}
\end{subfigure}%
\begin{subfigure}{0.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{logit_vs_loss}
\caption{\\Logit-Loss correlation}
\end{subfigure}
\caption{ Plots of various statistical measures: \textbf{(a)} Variations in the logit
norm vs. $\alpha$. The logit norm is calculated per image over the test set of
ImageNet-1K (at the linear layer of the ResNet-18 architecture) and then averaged.
\textbf{(b)} Variations in the final loss values obtained for different $\alpha$
settings. \textbf{(c)} Correlation between loss and logits. Regression line equation
:$(0.01074 x + 0.25816)$. }
\label{fig:stats}
\end{figure}

>>>>

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{framed}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{url}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{booktabs}
\usepackage{pifont}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{color}
\graphicspath{{images/}}
\renewcommand\UrlFont{\color{blue}\rmfamily}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\NewDocumentCommand{\codeword}{v}{\texttt{\textcolor{orange}{#1}}}
\newcommand{\cmark}{\ding{51}} \newcommand{\xmark}{\ding{55}}
\definecolor{newcolor}{rgb}{.8,.349,.1}
\definecolor{codegreen}{rgb}{0.7,0.7,0.7}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{1,1,1}
\lstdefinestyle{mystyle}{ backgroundcolor=\color{backcolour},   
commentstyle=\color{codegreen}, keywordstyle=\color{magenta},
numberstyle=\tiny\color{codegray}, stringstyle=\color{codepurple},
basicstyle=\ttfamily\footnotesize, breakatwhitespace=false,         
breaklines=true,                 
captionpos=b,                    
keepspaces=true,                 
numbersep=5pt,                  
showspaces=false,                
showstringspaces=false, showtabs=false,                  
tabsize=2 }
\lstset{style=mystyle}

\begin{document}
\title{Altering Backward Pass Gradients to improve Convergence}
\author{Bishshoy Das\inst{1}\orcidID{0000-0002-6525-4015}\and Milton Mondal\inst{1}\orcidID{xxxx-xxxx-xxxx-xxxx}\and Brejesh Lall\inst{3}\orcidID{xxxx-xxxx-xxxx-xxxx}\and Shiv Dutt Joshi\inst{3}\orcidID{0000-0002-0326-0524}\and Sumantra Dutta Roy\inst{3}\orcidID{0000-0002-2141-5067}}
\authorrunning{B. Das et al.}
\institute{Indian Institute of Technology Delhi, New Delhi - 110016, India \\ \email{\{bishshoy.das, milton.mondal, brejesh, sdjoshi, sumantra\}@ee.iitd.ac.in}}
\maketitle

@@

\subsection{Properties}
\label{sec:pgt_prop}

@@

@@

We have established that values of $p_i$ which are greater than the stationary threshold
decreases and move down towards the stationary threshold, and values in $p_i$ lower than
the stationary threshold moves up towards the stationary threshold. Therefore, this
transformation makes the distribution more uniform (i.e. it smooths out the actual
distribution) as $\alpha$ is decreased from $1$ and down towards $0$. This final
observation in the following theorem.

\subsection{Effects of PGT on logit, weight gradients and class separation}

PGT, other than having the properties mentioned above, also:
\begin{enumerate}
\item restricts the magnitude of the directional derivative for each logit from becoming too small or too high
\item enhances gradient behavior of weights in a manner that improves learning
\item leads to better class separation.
\end{enumerate}

We demonstrate each of these effects in detail in supplementary sections
\ref{sec:pgt_logit_derivative}, \ref{sec:pgt_weight_gradients} and
\ref{sec:pgt_class_separation} respectively.

\begin{table}[!t]
\centering
\caption{ Results and comparison table for CNN networks trained on ImageNet-1K (sorted
in descending order w.r.t. the last column denoting test accuracy improvements over
baseline). Best training and test accuracies are highlighted in red and blue
respectively. Accuracy differences are highlighted in yellow. In all experiments, we
observe that PGT positively improves both training and test accuracies. We also observe
that deeper networks such as SE-ResNet-50 and ResNet-50 accrue higher gains over the
baselines as compared to shallow networks like ResNet-18. }
\label{tab:imagenet_table}
\begin{tabular}{cccccccc}
\multirow{2}{*}{\textbf{Model}} & \multirow{2}{*}{\textbf{Scheduler}} &
\multirow{2}{*}{\textbf{Method}} & \textbf{PGT} & \textbf{Train}
& \textbf{Train} & \textbf{Test} & \textbf{Test} \\
& & & \textbf{($\alpha$)} & \textbf{Acc.(\%)} & \textbf{Diff(\%)} &
\textbf{Acc.(\%)} & \textbf{Diff(\%)} \\
\midrule
\multirow{2}{*}{SE-ResNet-50} & \multirow{2}{*}{Cosine} & Baseline & - & 81.5 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.97}}} & 77.218 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.734}}} \\
& & PGT & 0.3 & \textcolor{red}{\textbf{82.47}} & & \textcolor{blue}{\textbf{77.952}} &
\\
\midrule
\multirow{2}{*}{ResNet-50} & \multirow{2}{*}{Cosine} & Baseline & - & 79.18 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.5}}} & 76.56 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.656}}} \\
& & PGT & 0.05 & \textcolor{red}{\textbf{79.68}} & & \textcolor{blue}{\textbf{77.216}} &
\\
\midrule
\multirow{2}{*}{ResNet-50} & \multirow{2}{*}{Step} & Baseline & - & 78.99 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.57}}} & 75.97 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.524}}} \\
& & PGT & 0.05 & \textcolor{red}{\textbf{79.56}} & & \textcolor{blue}{\textbf{76.494}} &
\\
\midrule
\multirow{2}{*}{ResNet-101} & \multirow{2}{*}{Cosine} & Baseline & - & 82.29 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.81}}} & 77.896 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.362}}} \\
& & PGT & 0.3 & \textcolor{red}{\textbf{83.1}} & & \textcolor{blue}{\textbf{78.258}} &
\\
\midrule
\multirow{2}{*}{SE-ResNet-18} & \multirow{2}{*}{Cosine} & Baseline & - & 71.42 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.18}}} & 71.09 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.346}}} \\
& & PGT & 0.25 & \textcolor{red}{\textbf{71.6}} & & \textcolor{blue}{\textbf{71.436}} &
\\
\midrule
\multirow{2}{*}{ResNet-18} & \multirow{2}{*}{Step} & Baseline & - & 69.95 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.35}}} & 69.704 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.14}}} \\
& & PGT & 0.25 & \textcolor{red}{\textbf{70.3}} & & \textcolor{blue}{\textbf{69.844}} &
\\
\midrule
\multirow{2}{*}{ResNet-18} & \multirow{2}{*}{Cosine} & Baseline & - & 70.38 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.15}}} & 70.208 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.09}}} \\
& & PGT & 0.25 & \textcolor{red}{\textbf{70.53}} & & \textcolor{blue}{\textbf{70.298}} &
\\
\end{tabular}
\end{table}

In our experiments with Squeeze-and-Excitation variant of ResNet-50 i.e.
SE-ResNet-50\cite{hu2018squeeze}, we observe significant improvements; a $0.97\%$ boost
in training accuracy and a $0.734\%$ increase in test accuracy. In our experiments with
ResNet-50, we find an increase of $0.5\%$ and $0.656\%$ performance enhancement over the
cosine scheduler baseline for training and testing respectively. The corresponding
improvement over the step scheduler baseline for ResNet-50 is $0.57\%$ (training) and
$0.524\%$ (testing). ResNet-101 sees a higher improvement in training fit, $0.81\%$ to
be exact, while the improvement over the test set is $0.362\%$. Smaller networks such as
SE-ResNet-18 and ResNet-18 sees accuracy boosts which are smaller but nevertheless
positive. Therefore we conclude that with PGT, the consistent improvements in training
accuracies across all cases is because the networks train better and arrive at better
optimas during convergence. Per epoch training and test accuracy plots of ResNet-18 and
ResNet-50 (both with and without PGT) are shown in Fig. \ref{fig:metrics}(a, b). For
better visualization of the top end of the plots, we scale both axes in log scale.
Because both training and test accuracies improve, it indicates that PowerGrad Transform
allows the network to learn better representations\footnote{Reproducible code, training
recipes for the experiments, pretrained checkpoints and training logs are provided at:
\url{ https://github.com/skalien/power-grad-transform }. Code is adapted from
\cite{rw2019timm}.}.

We examine issues that occur in non-BatchNormalized networks and we provide insight into
the effects of PGT on such networks. We use ResNet-18 \cite{he2016deep} as the
foundation model trained on ImageNet-1K \cite{russakovsky2015imagenet}. We found deeper
networks (such as ResNet-34 and ResNet-50 models) are impossible to train without Batch
Normalization because of increased depth. So in order to conduct our empirical
experiments, we solely focus on the ResNet-18 architecture. We designate different
layers with their corresponding layer indices. We provide a detailed layer-wise index
list of ResNet-18 in supplementary section \ref{sec:pgt_class_separation}.

Throughout the training process, we monitor variations in the the per-filter L2-norm of
each layer's weight tensor. Because a full plot of each layer's data would use an
inordinate amount of space, we focus on a few critical layers' plots. The supplementary
sections (\ref{sec:plots1} through \ref{sec:plots4}) contains full plots for all layer
for various settings. The norm of the weights of the $11^\textit{th}$ convolutional
layer is shown in Fig. \ref{fig:norm_plots}(a). Layer $11$ includes $256$ filters. Each
colour on the figure represents the iteration-wise evolution of a particular filter
throughout the course of training. The critical point to note is that some filters
achieve a norm of zero during training. We refer to this event as `Zeroing Out', and
occurs when a channel of a weight tensor gets fully filled with zeros. If all the
coefficients of the filter becomes zero, only then the L2 norm of that filter becomes
zero. If we observe zero filter norm for a filter, then that filter does not contribute
at all to determine the input-output relationship of a dataset, as the feature tensor it
produces is also an all-zero tensor. According to our findings, when a filter tensor
becomes zeroed out, it does not recover with further training. We observe both effects
\textbf{1)} zeroed out filter \textbf{2)} zeroed out feature.

\begin{table}[!t]
\centering
\caption{ Results for non-normalized ResNets on ImageNet-1K. Each experiment is
conducted with a 100 epoch budget using the cosine decay scheduler. Best training and
test accuracies are highlighted in red and blue respectively. Top differences in
training and test accuracies are marked in yellow. The baselines in each segment is
taken as the reference point for all `Diff' accuracies in that segment. The first three
segments highlight comparison for experiments where the methods are applied
individually, while the last segment denotes performance metrics for combinations of
different methods (GC + PGT, AGC + PGT). We see that at any batch size, with PGT, the
performance of non-normalized version of ResNet-18 improves significantly over the
baseline owing to the improvements in training as observed in Fig. \ref{fig:agc_pgt}. }
\label{tab:wobn_table}
\begin{tabular}{cccccccc}
\multirow{2}{*}{\textbf{Model}} & \multirow{2}{*}{\textbf{Batch Size}} &
\multirow{2}{*}{\textbf{Method}} & \textbf{PGT} & \textbf{Train}
& \textbf{Train} & \textbf{Test} & \textbf{Test} \\
& & & \textbf{($\alpha$)} & \textbf{Acc.(\%)} & \textbf{Diff(\%)} &
\textbf{Acc.(\%)} & \textbf{Diff(\%)} \\
\midrule
ResNet-18 & 1024 & Baseline & - & 66.27 & - & 64.816 & - \\
(non-BN) & 1024 & PGT & 0.92 & \textcolor{red}{\textbf{66.62}} &
\textcolor{olive}{\textbf{+0.35}} & \textcolor{blue}{\textbf{65.498}} &
\textcolor{olive}{\textbf{+0.682}} \\
\cmidrule{2-8}
& 512 & Baseline & - & 68.02 & - & 66.552 & - \\
& 512 & PGT & 0.25 & \textcolor{red}{\textbf{69.5}} & \textcolor{olive}{\textbf{+1.48}}
& \textcolor{blue}{\textbf{67.236}} & \textcolor{olive}{\textbf{+0.684}} \\
\cmidrule{2-8}
& 256 & Baseline & - & 68.86 & - & 66.796 & - \\
& 256 & GC & - & 69.04 & +0.18 & 67.064 & +0.268 \\
& 256 & AGC & - & 69.06 & +0.2 & 67.298 & +0.502 \\
& 256 & PGT & 0.25 & \textcolor{red}{\textbf{69.97}} & \textcolor{olive}{\textbf{+1.11}}
& \textcolor{blue}{\textbf{67.814}} & \textcolor{olive}{\textbf{+1.018}} \\
\cmidrule{2-8}
& 256 & Baseline & - & 68.86 & - & 66.796 & - \\
& 256 & GC+PGT & 0.25 & 68.67 & -0.19 & 67.088 & +0.292 \\
& 256 & AGC+PGT & 0.25 & \textcolor{red}{\textbf{70.92}} &
\textcolor{olive}{\textbf{+2.06}} & \textcolor{blue}{\textbf{68.856}} &
\textcolor{olive}{\textbf{+2.06}} \\
\end{tabular}
\end{table}

\begin{figure}[t]
\centering
\begin{subfigure}{0.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{logit_vs_alpha}
\caption{\\Logit vs. $\alpha$}
\end{subfigure}%
\begin{subfigure}{0.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{loss_vs_alpha}
\caption{\\Cross-entropy loss vs. $\alpha$}
\end{subfigure}%
\begin{subfigure}{0.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{logit_vs_loss}
\caption{\\Logit-Loss correlation}
\end{subfigure}
\caption{ Plots of various statistical measures: \textbf{(a)} Variations in the logit
norm vs. $\alpha$. The logit norm is calculated per image over the test set of
ImageNet-1K (at the linear layer of the ResNet-18 architecture) and then averaged.
\textbf{(b)} Variations in the final loss values obtained for different $\alpha$
settings. \textbf{(c)} Correlation between loss and logits. Regression line equation:
$(0.01074 x + 0.25816)$. }
\label{fig:stats}
\end{figure}

\begin{table}[!t]
\centering
\caption{ Ablation study for ResNet-50 on ImageNet-1K. }
\label{tab:ablation_table}
\begin{tabular}{cccccccc}
\multirow{2}{*}{\textbf{\#Row}} & \multirow{2}{*}{\textbf{Model}} &
\multirow{2}{*}{\textbf{Scheduler}} & \textbf{Label} & \textbf{PGT} &
\textbf{Train} & \textbf{Test} & \multirow{2}{*}{\textbf{Gap(\%)}} \\
& & & \textbf{Smoothing} & \textbf{($\alpha$)} & \textbf{Acc.(\%)} &
\textbf{Acc.(\%)} & \\
\midrule
1. & ResNet-50 & Step & \xmark & \xmark & 78.99 & 75.97 & 3.02 \\
2. & ResNet-50 & Step & \xmark & 0.3 & 79.56 & 76.494 & 3.066 \\
\cmidrule{2-8}
3. & ResNet-50 & Cosine & \xmark & \xmark & 79.18 & 76.56 & 2.62 \\
4. & ResNet-50 & Cosine & 0.1 & \xmark & 78.81 & 76.698 & 2.112 \\
\cmidrule{2-8}
5. & ResNet-50 & Cosine & \xmark & 0.3 & 79.43 & 76.886 & 2.544 \\
6. & ResNet-50 & Cosine & 0.1 & 0.3 & 78.47 & 76.968 & 1.502 \\
\cmidrule{2-8}
7. & ResNet-50 & Cosine & \xmark & 0.05 & 79.68 & 77.216 & 2.464 \\
8. & ResNet-50 & Cosine & 0.1 & 0.05 & 77.69 & 76.39 & 1.3 \\
\end{tabular}
\end{table}

% \end{document}

\subsection{Proofs}

@@

@@

If $a_1,\dots ,a_n$ and $b_1,\dots ,b_n$ are non-negative numbers, then using the log
sum inequality, \\ we get $\sum _{j=1}^n a_j \log \left(\frac{a_j}{b_j}\right)\geq
\left(\sum _{j=1}^n a_j\right) \log \left(\frac{\sum _{j=1}^n a_j}{\sum _{j=1}^n
b_j}\right)$.

Setting $a_j=p_j^\alpha$ and $b_j=1$, we get the following lower bound
\begin{equation} \sum _{j=1}^C p_j^{\alpha } \log
\left(p_j^{\alpha }\right)\geq \left(\sum _{j=1}^C
p_j^{\alpha}\right) \log \left(\frac{1}{C} \sum _{j=1}^C
p_j^{\alpha}\right) \label{lower_bound1} \end{equation}

@@

\begin{equation} \frac{\partial}{\partial \alpha
}\left(\sum _{j=1}^c p_j^{\alpha}\right){}^{\frac{1}{\alpha -1}}\ \ \geq \ \ 0
\end{equation}

@@

Now we demonstrate how PGT restricts the magnitude of the directional derivative for
each logit. For a training example, a neural network can assign the highest probability
to either (i) a wrong class or to the (ii) correct class.

@@
